---
title: "mutenm package"
author: "Julian Echave"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  message = FALSE,
  warning = FALSE
)
```

The `mutenm` package builds Elastic Network Models (ENMs) of proteins and
simulates mutation effects using linear response theory.

# Installation

```
# install.packages("devtools")
devtools::install_github("jechave/mutenm")
```

# Overview

## What is an ENM?

An Elastic Network Model represents a protein as a network of nodes (typically
Cα atoms) connected by harmonic springs. The model captures the protein's
intrinsic flexibility through normal mode analysis, producing a covariance
matrix that describes thermal fluctuations.

## How are mutations simulated?

Mutations are simulated by perturbing the equilibrium lengths of springs
connected to the mutated site. The structural response is calculated using
linear response theory:

```
δr = C × f
```

where `C` is the covariance matrix and `f` is the force from edge perturbations.

Two mutation models are available:

- **lfenm**: Fast. Updates structure only; dynamics unchanged.
- **sclfenm**: Slower. Recalculates the full ENM, capturing changes in dynamics
  and entropy.

## What can you calculate?

- **Structure**: contact number, weighted contact number, distance to active site
- **Dynamics**: mean-square fluctuations per site or per mode
- **Energy**: minimum (stress) energy, entropic free energy, activation energies
- **Mutation effects**: changes in structure, dynamics, and energy between
  wild-type and mutant

# Key Functions

## Core

| Function | Description |
|----------|-------------|
| `enm(pdb, node, model, d_max)` | Build an ENM from a PDB structure |
| `mutenm(wt, site_mut, ...)` | Simulate a mutation at a specific site |
| `mrs(wt, nmut, ...)` | Mutation Response Scanning — scan all sites |

## Single-protein properties

| Function | Description |
|----------|-------------|
| `cn(prot)` | Contact number per site |
| `wcn(prot)` | Weighted contact number per site |
| `dactive(prot, pdb_site_active)` | Distance to active site |
| `msfi(prot)` | Mean-square fluctuation per site |
| `msfn(prot)` | Mean-square fluctuation per mode |
| `v_min(prot)` | Minimum (stress) energy |
| `g_ent(prot, beta)` | Entropic free energy |
| `dv_act(prot, ideal, pdb_site_active)` | Activation energy (internal) |
| `dg_ent_act(prot, ideal, pdb_site_active, beta)` | Activation energy (entropic) |

## Mutation effects (comparing wild-type and mutant)

| Function | Description |
|----------|-------------|
| `Dr2i(wt, mut)` | Squared displacement per site |
| `Dr2n(wt, mut)` | Squared displacement per mode |
| `Dmsfi(wt, mut)` | Change in MSF per site (requires sclfenm) |
| `Dmsfn(wt, mut)` | Change in MSF per mode (requires sclfenm) |
| `Dv_min(wt, mut)` | Change in minimum energy |
| `Dg_ent(wt, mut, beta)` | Change in entropic free energy |
| `Ddv_act(wt, mut, ...)` | Change in activation energy (internal) |
| `Ddg_ent_act(wt, mut, ...)` | Change in activation energy (entropic) |

# Quick Start

```{r}
library(mutenm)
```

## 1. Build an ENM

Read a PDB file and create the elastic network model:

```{r, results = 'hide'}
pdb <- bio3d::read.pdb("data-raw/2acy.pdb")
wt <- enm(pdb, node = "ca", model = "ming_wall", d_max = 10.5)
```

The `wt` object contains the protein structure, network topology, and normal modes.

Calculate structural and dynamic properties:

```{r}
# Contact number per site
head(cn(wt))

# Mean-square fluctuation per site
head(msfi(wt))

# Distance to active site (residues 23 and 41 for 2acy)
head(dactive(wt, pdb_site_active = c(23, 41)))
```

## 2. Simulate a mutation

Create a mutant by perturbing contacts at a specific site:

```{r}
mut <- mutenm(wt, site_mut = 50, mutation = 1, mut_model = "lfenm", seed = 123)
```

## 3. Analyze mutation effects

Compare wild-type and mutant properties:

```{r}
# Energy change (stress energy)
dv <- Dv_min(wt, mut)
dv

# Structural displacement at each site
dr2 <- Dr2i(wt, mut)
head(dr2)
```

## 4. Mutation Response Scanning

Scan all sites systematically with `mrs()`:

```{r}
# Scan all sites with 3 mutations each
responses <- mrs(wt, nmut = 3, mut_model = "lfenm", seed = 42)

# Response tibbles
names(responses)

# Energy change per mutation site
head(responses$Dv_min)

# Structural response: displacement at site i due to mutation at site j
head(responses$Dr2i)
```

See `?mutenm` for the full package documentation.
