---
title: "mutenm package"
author: "Julian Echave"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  message = FALSE,
  warning = FALSE
)
```

The `mutenm` package contains functions to build Elastic Network Models (ENM) of
proteins and to perturb them;
`mutenm` stands for Mutating Elastic Network Models.


# Installation

Install packages `mutenm` (this package) and `jefuns` (miscellaneous functions, some of which `mutenm` uses).

```
# install.packages("devtools")

devtools::install_github("jechave/jefuns")
devtools::install_github("jechave/mutenm")
```

# Quick Start

```{r}
library(mutenm)
```

## Set up the ENM for a protein

First, read a pdb file using `bio3d::read.pdb` to generate a `pdb` object for a protein.
Then, create the `prot` object, that contains the full ENM analysis.

```{r, results = 'hide'}
pdb <- bio3d::read.pdb("data-raw/2XWRa.pdb") # read a pdb file
wt <- enm(pdb, node = "calpha", model = "anm", d_max = 10.5)
```

`wt` created here by `enm()` is an object of class _prot_.
In this example, network nodes are placed at $C_\alpha$ coordinates, the model used is Bahar's Anisotropic Network Model (`model = "anm"`) with a cut-off distance to define contacts of `d_max = 10.5`.

## Create a mutant

Use `mutenm()` to create a mutant at a specific site:

```{r}
mut <- mutenm(wt, site_mut = 50, mutation = 1, mut_model = "lfenm", mut_dl_sigma = 0.3, seed = 123)
```

This creates a mutant protein by perturbing contacts at site 50 using the LFENM model.

## Compare wild-type and mutant

Calculate the stability change (ΔΔG) between wild-type and mutant:

```{r}
ddg <- ddg_dv(wt, mut)
ddg
```

See the package documentation for more detailed examples and mutation scanning functions.
